---
title: "Untitled"
author: "Jennifer Kampe"
date: "6/15/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages

```{r, echo=FALSE, message = FALSE, warning = FALSE}
# Load packages 
library(Hmsc)
library(tidyverse)
library(magrittr)
library(tibble)
library(igraph)
library(zoo)
library(geobr)
library(abind)
library(vioplot)
library(colorspace)
library(knitr)
```

Load files

```{r}
frug <- read.csv("./Data/Frugivory/frug_clean.csv", header = TRUE)
birds <- read.csv("./Data/atlantic_birds/birds_wide_genus_sp_covars.csv", header = TRUE)
trees <- read.csv("./Data/Trees/trees_wide_covars.csv", header = TRUE)


### SPECIES OCCURRENCE: select just the most common trees and convert to binary
trees_occ <- trees[, which(colnames(trees)=="Abarema.cochliacarpos"):which(colnames(trees) =="Zygia.latifolia")]
trees_occ <- ifelse(trees_occ > 0, 1, 0)
ntrees <- 25
commontrees <- colnames(trees_occ[, order(colSums(-trees_occ))])[1:ntrees]

birds.df <- birds %>% 
  mutate(Habitat2 = ifelse(Olsoneconame == "Alto Parana Atlantic forests" 
                           | Olsoneconame == "Xingu-Tocantins-Araguaia moist forests", "Tropical Moist Forest", 
                           ifelse(str_detect(birds$Olsoneconame, "interior"), "Interior Forest", 
                                  ifelse(Olsoneconame == "Cerrado" | str_detect(birds$Olsoneconame, "savanna"), "Savanna", 
                                         ifelse(str_detect(birds$Olsoneconame, "coastal")
                                                | Olsoneconame == "Southern Atlantic mangroves", "Coastal Forest", 
                                                ifelse(Olsoneconame =="Araucaria moist forests", "Temperate Forest", NA
                                                       #ifelse(Olsoneconame == "Caatinga", "Shrubland", NA)
                                                )))))) %>%
  select(-c(Methods, Year_start, Year_finish, 
            Habitat, State, replicate, year, name_biome, code_biome)) %>% 
  relocate(Habitat2, .after = "ID_codref") %>%
  na.omit


## Species Occurrence: Y Select just the most common birds
birds_occ <- birds.df[, which(colnames(birds.df)=="Aburria.jacutinga"):which(colnames(birds.df) == "Zonotrichia.capensis")]
nbirds <- 10
commonbirds <- colnames(birds_occ[, order(colSums(-birds_occ))])[1:nbirds]

# Get adjacency matrix
frug.common <- frug %>%
  mutate(Bird_species = gsub(" ", ".", Bird_species),
         Tree_species = gsub(" ", ".", Tree_species)) %>% 
  filter(Bird_species %in% commonbirds, Tree_species %in% commontrees)

```

Look at interaction data. Should restrict to those being used in model
```{r}
el <- as.data.frame(frug.common[, c(2,3)])
t.el <- table(el)
g <- graph.incidence(t.el, weighted = TRUE)
LO <- layout_as_bipartite(g)
colrs <- c("light blue", "coral")[V(g)$type + 1L]
plot(g, vertex.color = colrs, layout = LO[, 2:1], vertex.label.cex = 0.75)

```

# Model 1: 10 birds, predicted using 25 trees using frugivory-informed priors
```{r}

# Look at tree fit
filename0 = "Models/mod1trees_HMSC.Rdata"
load(filename0)
# MCMC diagnostics
m1posttrees <- convertToCodaObject(mod1trees_HMSC)
hist(effectiveSize(m1posttrees$Beta), main="ess(beta)")
hist(gelman.diag(m1posttrees$Beta, multivariate=FALSE)$psrf, main="psrf(beta)")

# Look at params
postBeta = getPostEstimate(mod1trees_HMSC, parName = "Beta")
plotBeta(mod1trees_HMSC, post = postBeta, param = "Support", supportLevel = 0.95)

# Look at fit
preds.spatial.trees1 = computePredictedValues(mod1trees_HMSC)
MF.spatial.trees1= evaluateModelFit(hM=mod1trees_HMSC, predY=preds.spatial.trees1)
MF.spatial.trees1

filename1 = "Models/mod1_HMSC.Rdata"
load(filename1)

# MCMC diagnostics
m1post <- convertToCodaObject(mod1_HMSC)
hist(effectiveSize(m1post$Beta), main="ess(beta)")
hist(gelman.diag(m1post$Beta, multivariate=FALSE)$psrf, main="psrf(beta)")

# Look at params
postBeta = getPostEstimate(mod1_HMSC, parName = "Beta")
plotBeta(mod1_HMSC, post = postBeta, param = "Support", supportLevel = 0.95)

# Look at fit
preds.spatial.1 = computePredictedValues(mod1_HMSC)
MF.spatial.1= evaluateModelFit(hM=mod1_HMSC, predY=preds.spatial.1)


filename2 = "Models/mod2_HMSC.Rdata"
load(filename2)
# Look at fit
preds.spatial.2 = computePredictedValues(mod1_HMSC)
MF.spatial.2= evaluateModelFit(hM=mod1_HMSC, predY=preds.spatial.2)


filename3 = "Models/mod3_HMSC.Rdata"
load(filename3)
# Look at fit
preds.spatial.3 = computePredictedValues(mod3_HMSC)
MF.spatial.3= evaluateModelFit(hM=mod3_HMSC, predY=preds.spatial.3)

```

# Compare Fit:RMSE
```{r}
df.rmse <- as.data.frame(rbind(MF.spatial.1$RMSE, MF.spatial.2$RMSE, MF.spatial.3$RMSE))
row.names(df.rmse) <- c("Trees+", "Trees", "No Trees")
colnames(df.rmse) <- commonbirds
kable(df.rmse)
```

# Compare Fit: AUC
```{r}
df.auc <- as.data.frame(rbind(MF.spatial.1$AUC, MF.spatial.2$AUC, MF.spatial.3$AUC))
row.names(df.auc) <- c("Trees+", "Trees", "No Trees")
colnames(df.auc) <- commonbirds
kable(df.auc)
```

Modifications / To Do

- Look at the quality of tree predictions at held out locations not immediately next to observed locations
- Modify included plants and trees to be common in the frugivory dataset rather than common in the disctinct bird, tree sets
- Run longer
- Replace variable selection prior probs with Georgia posterior estimates
- Add globally available climate and landuse data 
- Indirect interactions?
- Negative interactions - habitat proxy? - inclue tree PCs to cover habitat
